---
title: 'UVA Baseball: Pitching Optimization'
author: "Adam Chow"
date: "2024-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## LOAD PACKAGES HERE
library(tidyverse)
library(readxl)
library(stringr)
library(rvest)
library(plotly)
library(ggplot2)
library(shiny)

war_values <- c(
  3.1,-1.1,4.2,0.7,2.6,
  0.2,1.0,1.7,2.0,3.3,
  1.7,2.1,2.2,0.3,-1.2,
  2.0,3.1,0.6,-0.5,2.2,
  2.0,3.9,1.7,2.3,1.8,
  5.4,-0.4,3.0,1.4,2.4,
  1.8,0.1,0.8,2.8,2.8,
  2.4,1.7,0.1,0.6,1.5,
  5.5,1.1,1.2,-1.8,1.1,
  2.9,2.3,-0.5,2.8,2.7,
  0.5,3.2,1.9,0,1.8,
  1.8,1.1,2.5,3.7,0,
  1.8,4.5,-1.4,2.4,0.7,
  2.2,0.3,2.5,0.7,-1.3,
  0.8,2.1,1.5,3.5,1.8,
  1.7,2.9,0.8,0.4,1.6,
  2.0,1.0,0.9,0.8,1.8,
  1.3,3.9,2.2,1.3,1.5,
  1.0,2.1,5.1,1.6,1.3,
  3.2,4.6,5.4,2.1,1.2,
  2.1,2.3,1.5,2.3,3.3,
  2.4,1.1,2.8,2.6,1.0,
  3.6,4.4,0.7,2.1
)

left <- c(
  1,1,1,0,0,
  0,0,0,0,0,
  0,0,0,0,0,
  0,0,0,1,1,
  0,1,0,0,1,
  0,0,0,0,1,
  0,0,0,0,0,
  0,1,1,0,0,
  0,1,1,0,0,
  0,1,1,0,0,
  1,0,0,0,0,
  0,0,0,0,0,
  0,0,0,1,0,
  0,0,0,0,1,
  0,0,0,0,0,
  0,0,1,1,0,
  0,0,0,1,0,
  1,1,0,0,1,
  1,0,1,1,0,
  0,0,1,0,0,
  1,0,0,0,0,
  1,0,1,0,0,
  0,0,0,0
)
```

# Background

## Import and Clean Data
```{r}
# Import 2024 MLB Pitching Data
setwd("/Users/adamchow/Library/CloudStorage/Box-Box/UVA/Fall 2024 Classes/STAT 4996/")

pitches <- read_csv("mlb_pitch_data.csv")
expected <- read_csv("mlb_exp_stats.csv")

mlb2024_df <- left_join(expected, pitches, by = "player_id")

mlb2024_df <- mlb2024_df %>%
  select(-pa, -player_id, -year.x, -year.y, -bip, -ba, -est_ba,
         -est_ba_minus_ba_diff, -slg, -est_slg,
         -est_slg_minus_slg_diff, -`last_name, first_name.y`) %>%
  rename(name = `last_name, first_name.x`,
         ev_avg = `exit_velocity_avg`,
         fb_percent = `n_ff_formatted`,
         fb_avg_velo = `ff_avg_speed`,
         fb_avg_spin = `ff_avg_spin`,
         fb_avg_hor = `ff_avg_break_x`,
         fb_avg_vert = `ff_avg_break_z`,
         sl_percent = `n_sl_formatted`,
         sl_avg_velo = `sl_avg_speed`,
         sl_avg_spin = `sl_avg_spin`,
         sl_avg_hor = `sl_avg_break_x`,
         sl_avg_vert = `sl_avg_break_z`,
         ch_percent = `n_ch_formatted`,
         ch_avg_velo = `ch_avg_speed`,
         ch_avg_spin = `ch_avg_spin`,
         ch_avg_hor = `ch_avg_break_x`,
         ch_avg_vert = `ch_avg_break_z`,
         cb_percent = `n_cu_formatted`,
         cb_avg_velo = `cu_avg_speed`,
         cb_avg_spin = `cu_avg_spin`,
         cb_avg_hor = `cu_avg_break_x`,
         cb_avg_vert = `cu_avg_break_z`,
         sk_percent = `n_si_formatted`,
         sk_avg_velo = `si_avg_speed`,
         sk_avg_spin = `si_avg_spin`,
         sk_avg_hor = `si_avg_break_x`,
         sk_avg_vert = `si_avg_break_z`,
         cu_percent = `n_fc_formatted`,
         cu_avg_velo = `fc_avg_speed`,
         cu_avg_spin = `fc_avg_spin`,
         cu_avg_hor = `fc_avg_break_x`,
         cu_avg_vert = `fc_avg_break_z`,
         sp_percent = `n_fs_formatted`,
         sp_avg_velo = `fs_avg_speed`,
         sp_avg_spin = `fs_avg_spin`,
         sp_avg_hor = `fs_avg_break_x`,
         sp_avg_vert = `fs_avg_break_z`,
         sw_percent = `n_st_formatted`,
         sw_avg_velo = `st_avg_speed`,
         sw_avg_spin = `st_avg_spin`,
         sw_avg_hor = `st_avg_break_x`,
         sw_avg_vert = `st_avg_break_z`) %>%
  mutate(war = NA,
         is_left = NA) %>%
  select(name, is_left, war, everything()) %>%
  slice(1:114) %>%
  arrange(name)

mlb2024_df$war[1:length(war_values)] <- war_values
mlb2024_df$is_left[1:length(left)] <- left
```


## Find Most Effective Pitchers
```{r}
pitcher_success <- mlb2024_df %>%
  select(1, 3, 4, 5, 7, 8, 10) %>%
  column_to_rownames(var = "name") %>%
  scale()

clusters<-10

##store between and within cluster variance
within<-between<-array(0,clusters)

for (k in 2:clusters)
  
{
  set.seed(2)
  km.out<-kmeans(pitcher_success, centers=k, nstart=20)
  
  between[k]<-km.out$betweenss
  within[k]<-km.out$tot.withinss
  
}

plot(c(2:clusters), between[-1], xlab="K", ylab="Between Cluster Variance", type='b')
plot(c(2:clusters), within[-1], xlab="K", ylab="Total Within Cluster Variance", type='b')
```

By this plots we think that 4 clusters is the most appropriate number given the plot.

### Create a 2D Plot of CLusters
```{r}
##4 means clustering
set.seed(2)
km.out4<-kmeans(pitcher_success, centers=4, nstart=20)

##note UA.S after scaling is not a dataframe, so we convert it to one
pitcher_success<-data.frame(pitcher_success)

##scatterplot with different colors based on cluster assignment from 4-means 
plot(pitcher_success$war,pitcher_success$era, pch=20, xlab="WAR", ylab="ERA", col=(km.out4$cluster+1))
text(pitcher_success$war,pitcher_success$era, labels=rownames(pitcher_success), cex=0.75, font=1, col=(km.out4$cluster+1))
```

```{r}
km.out4
```

Out of the 114 Qualified Starting Pitchers in 2024, here is a list of the most elite pitchers this season:
1. Corbin Burnes (R) 6-3, 245lbs
2. Hunter Brown (R) 6-2, 220lbs
3. Dylan Cease (R) 6-2, 200lbs
4. Garrett Crochet (L) 6-6, 245lbs
5. Jack Flaherty (R) 6-4, 225lbs
6. Max Fried (L) 6-4, 190lbs
7. Logan Gilbert (R) 6-6, 215lbs
8. Tyler Glasnow (R) 6-8, 225lbs
9. Hunter Greene (R) 6-5, 242lbs
10. Michael King (R) 6-3, 210lbs
11. Nick Martinez (R) 6-1, 200lbs
12. Cole Ragans (L) 6-4, 190lbs
13. Joe Ryan (R) 6-2, 205lbs
14. Chris Sale (L) 6-6, 180lbs
15. Paul Skenes (R) 6-6, 235lbs
16. Tarik Skubal (L) 6-3, 240lbs
17. Justin Steele (L) 6-2, 205lbs
18. Ranger Suarez (R) 6-1, 217lbs
19. Zack Wheeler (R) 6-4, 195lbs

## Identify qualities that make pitchers elite
```{r}
# Find the Appropriate number of Clusters for these elite pitchers
last_names <- c("Burnes", "Brown", "Cease", "Crochet", "Flaherty", 
                "Fried", "Gilbert", "Glasnow", "Greene", "King", 
                "Martinez", "Ragans", "Ryan", "Sale", "Skenes", 
                "Skubal", "Steele", "Wheeler")

# Extract the last names from the "name" column
mlb2024_df$last_name <- str_extract(mlb2024_df$name, "^[^,]+")

# Subset the dataframe based on the last names
elite_pitchers <- mlb2024_df %>%
  filter(last_name %in% last_names | row_number() == 105) %>%
  select(-last_name) %>%
  mutate(
    height = c(75, 74, 74, 78, 76, 76, 78, 80, 77, 75, 73, 76, 74, 78, 78, 75, 74, 73, 76),
    weight = c(245, 220, 200, 245, 225, 190, 215, 225, 242, 210, 200, 190, 205, 180, 235, 240, 205, 217, 195)
  )

# Find optimal Number of clusters depending on physical stature
elite_scale <- elite_pitchers %>%
  select(1, 51, 52) %>%
  column_to_rownames(var = "name") %>%
  scale()

clusters<-10

##store between and within cluster variance
within<-between<-array(0,clusters)

for (k in 2:clusters)
  
{
  set.seed(2)
  km.out<-kmeans(pitcher_success, centers=k, nstart=20)
  
  between[k]<-km.out$betweenss
  within[k]<-km.out$tot.withinss
  
}

plot(c(2:clusters), between[-1], xlab="K", ylab="Between Cluster Variance", type='b')
plot(c(2:clusters), within[-1], xlab="K", ylab="Total Within Cluster Variance", type='b')
```

By these plots we think that 4 clusters is the most appropriate number given the plot.

### Create a 2D Plot of CLusters
```{r}
##4 means clustering
set.seed(2024)
km.out4<-kmeans(elite_scale, centers=4, nstart=20)

##note UA.S after scaling is not a dataframe, so we convert it to one
elite_scale<-data.frame(elite_scale)

##scatterplot with different colors based on cluster assignment from 4-means 
plot(elite_scale$height,elite_scale$weight, pch=20, xlab="Height", ylab="Weight", col=(km.out4$cluster+1))
text(elite_scale$height,elite_scale$weight, labels=rownames(elite_scale), cex=0.75, font=1, col=(km.out4$cluster+1))
```

```{r}
km.out4
```

Cluster 1: Average Height (6'2"-6'5"), but Stocky build with emphasis on muscle to produce power
Cluster 2: Extremely Tall (6'6"-6'8") genetic anomalies, Also weight more than the average pitcher due to sheer height
Cluster 3: Average Height (6'2"-6'5"), but Lanky, slender build with emphasis on leverage and flexibility to produce power
Cluster 4: Undersized both in Height and Weight, most likely outsmarting hitters due to lack of overwhelming physical gifts (Might have to dive in deeper to see what each pitcher does differently to sustain effectiveness)

## Cluster 1
```{r}
elite_pitches <- elite_pitchers %>%
  select(-3, -4, -5, -6, -7, -8, -9, -10) %>%
  column_to_rownames(var = "name")

# Convert Vertical Break to IVB
elite_pitches <- elite_pitches %>%
  mutate(fb_avg_vert = fb_avg_vert + (523 / fb_avg_velo)^2) %>%
  mutate(sl_avg_vert = sl_avg_vert + (523 / sl_avg_velo)^2) %>%
  mutate(ch_avg_vert = ch_avg_vert + (523 / ch_avg_velo)^2) %>%
  mutate(cb_avg_vert = cb_avg_vert + (523 / cb_avg_velo)^2) %>%
  mutate(sk_avg_vert = sk_avg_vert + (523 / sk_avg_velo)^2) %>%
  mutate(cu_avg_vert = cu_avg_vert + (523 / cu_avg_velo)^2) %>%
  mutate(sp_avg_vert = sp_avg_vert + (523 / sp_avg_velo)^2) %>%
  mutate(sw_avg_vert = sw_avg_vert + (523 / sw_avg_velo)^2)

clusters <- c(
  "Brown, Hunter" = 1, "Burnes, Corbin" = 4, "Cease, Dylan" = 4, 
  "Crochet, Garrett" = 2, "Flaherty, Jack" = 1, "Fried, Max" = 3, 
  "Gilbert, Logan" = 2, "Glasnow, Tyler" = 2, "Greene, Hunter" = 1, 
  "King, Michael" = 4, "Martinez, Nick" = 4, "Ragans, Cole" = 3, 
  "Ryan, Joe" = 4, "Sale, Chris" = 3, "Skenes, Paul" = 2, 
  "Skubal, Tarik" = 1, "Steele, Justin" = 4, "Suárez, Ranger" = 4, 
  "Wheeler, Zack" = 3
)

# Split the elite_pitches dataframe into four separate dataframes
c1 <- elite_pitches[rownames(elite_pitches) %in% names(groupings[groupings == 1]), ]
c2 <- elite_pitches[rownames(elite_pitches) %in% names(groupings[groupings == 2]), ]
c3 <- elite_pitches[rownames(elite_pitches) %in% names(groupings[groupings == 3]), ]
c4 <- elite_pitches[rownames(elite_pitches) %in% names(groupings[groupings == 4]), ]
```

Cluster 1 (Stocky) Standard Pitch Arsenal: FB, SL, CH, CB, ~SK, ~CU
Cluster 2 (Genetic Freak) Standard Pitch Arsenal: FB, SL/SW, SK, ~CH/CB, ~CU
Cluster 3 (Lanky) Standard Pitch Arsenal: FB, SL/SW, CH/SP, ~CB, ~SK, ~CU
Cluster 4 (Anomalies) Standard Pitch Arsenal: Much more Variability SK-SL/CH is most used combination

*We want our UVA Pitchers to ideally fit into one of these Archetypes*

## Shiny Plotly Graph

**Shiny App**: 


# UVA Game Data
```{r}
library(tidyverse)
library(lubridate)

setwd("/Users/adamchow/Library/CloudStorage/Box-Box/UVA/Fall 2024 Classes/STAT 4996/Old Files")

process_game <- function(file) {
  read_csv(file) %>%
    select(2:70) %>%
    select(-22, -27) %>%
    mutate(Date = ymd(Date))
}

game1 <- process_game("game1.csv")
game2 <- process_game("game2.csv")
game3 <- process_game("game3.csv")
game4 <- process_game("game4.csv")
game5 <- process_game("game5.csv")
game6 <- process_game("game6.csv")
game7 <- process_game("game7.csv")
game8 <- process_game("game8.csv")

uva_games <- bind_rows(game1, game2, game3, game4, game5, game6, game7, game8) %>%
  filter(PitcherTeam == "VIR_CAV")

write_csv(uva_games, "filtered_uva_games.csv")
```

```{r}
heat <- game %>% 
      filter(Pitcher == "Edgington, Brian") %>%
      group_by(Pitcher,Count, Zone, TaggedPitchType, BatterSide) %>%
      summarise(
        total = n(),
        putaway = sum(PitchCall %in% c("StrikeCalled", "StrikeSwinging")),
        putaway_rate = round(putaway / total, 2),
        # Hard Hit Balls
        hhb = sum(ExitSpeed > 87),
        # In Play Avg EV
        in_ev = mean(ExitSpeed),
        .groups = 'drop'
      )

heat
```

```{r}
tab <- game %>%
      filter(Pitcher == "Edgington, Brian") %>%
      filter(Count == "0-0", Zone == 1)

tab
```












